Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-19 VMCAI-revised-version/final.tex
effectivity

Surprisingly, the general undecidability statement regarding resilience had neither been known nor proven.
We show that the resilience problem is undecidable for $\omega^2$-WSTS with strong compatibility.

\begin{proposition}\label{indec WSTS}
Resilience for $\post$-effective $\omega^2$-WSTS with strong compatibility, for upward closed sets $\Safe$ is undecidable.
\end{proposition}
% \alain{peut-on ajouter des hypothèses d'effectivity ? peut-on montrer que l'exemple n'est pas pred effectif ? l'exemple est-il  ideal-effective (oui pour les propriétes 1 et 2 mais pas 3 car ça permettrait de décider l'arrêt ?)? on voit que l'exemple n'est pas completion-$\post$-effective vu le theoreme qui suit}
\begin{proof}
Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
increasing recursive functions from \cite{FMP-ic04} defined as

@@

We start with the case $\Safe=\uparrow \Safe$, hence $\Bad=\downarrow \Bad$.
Since Resilience is undecidable for  $\omega^2$-WSTS with strong compatibility (Theorem \ref{indec WSTS}), we still consider $\omega^2$-WSTS with strong compatibility but by strengthening the assumptions of effectiveness (we now consider the $\post$-ideal-effective hypothesis) and we demonstrate that the three resilience problems are now decidable.

%
\begin{theorem}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a $\post$-ideal-effective WSTS with strong compatibility and a set $\Safe = \uparrow \Safe$.
{\sc Resilience}, {\sc Bounded resilience} 
and {\sc $k$-resilience} are decidable.
\end{theorem}

@@

\begin{proof}
Let us first recall two results in \cite{BFM-ic17}. Proposition 30 establishes a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ and the runs of its completion $\hat{\mathscr{S}}$. It states that if $x \xrightarrow{k} y$ in $\mathscr{S}$ then for every ideal $I \supseteq \downarrow x$, there exists an ideal $J \supseteq \downarrow y$ such that $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$. Proposition 29 establishes that if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ then for every $y \in J$, there exists $x \in I$ and $y' \geq y$ such that $x \xrightarrow{k'} y'$ in $\mathscr{S}$. Moreover, if $\mathscr{S}$ has transitive compatibility then $k’ \geq k$; if $\mathscr{S}$ has strong compatibility then $k’ = k$.
%

@@

%
\begin{theorem}\label{up-down}
{\sc Resilience}  is decidable for $\pred^*$-effective downward-WSTS with 
$\Safe=\downarrow \Safe$.
\end{theorem}

\begin{proof}
Let $\mathscr{S}$ be a $\pred^*$-effective downward-WSTS with 
$\Safe=\downarrow \Safe$.
By Lemma \ref{downward compatible}, $\pred^*(\Safe)$ is downward-closed; moreover, $\pred^*(\Safe)$ is computable because $\mathscr{S}$ is $\pred^*$-effective.
The resilience problem can be reformulated as $\Bad \cap (S \setminus \pred^*(\Safe)) = \emptyset$.
%\alain{mais comment je calcule $\pred^*(\Safe)$ ?}
Since $\leq$ is ideally-effective, we can compute complementaries and intersections of upward and downward-closed
 subsets.
Hence we can compute $S \setminus \pred^*(\Safe)$ and then the intersection of
$\Bad$
and
$S \setminus \pred^*(\Safe)$,
which are both upward-closed. We may decide whether this intersection is empty, hence
the resilience problem is decidable. \qed
\end{proof}

In the case
of a $\pred^*$-effective WSTS, not necesarly downward compatible,
the above construction
can provide a proof
of non-resilience
i.e. when
$\Bad \cap (S \setminus \downarrow\pred^*(\Safe)) \neq \emptyset$
then
$\Bad \not\subseteq \downarrow\pred^*(\Safe)$
and hence
$\Bad \not\subseteq \pred^*(\Safe)$.
When $\Bad \cap (S \setminus \downarrow\pred^*(\Safe)) = \emptyset$
however
it is not enough to conclude.\\

>>>>

Surprisingly, the general undecidability statement regarding resilience had neither been known nor proven.
We show that the resilience problem is undecidable for $\omega^2$-WSTS with strong compatibility. 
%
\alain{pourquoi n'a-t-on pas prouvé la même chose pour la BR et la k-R ?: où est la difficulté ?}
\begin{proposition}\label{indec WSTS}
Resilience for $\post$-effective $\omega^2$-WSTS with strong compatibility, for upward closed sets $\Safe$ is undecidable.
\end{proposition}
% \alain{peut-on ajouter des hypothèses d'effectivity ? peut-on montrer que l'exemple n'est pas pred effectif ? l'exemple est-il  ideal-effective (oui pour les propriétes 1 et 2 mais pas 3 car ça permettrait de décider l'arrêt ?)? on voit que l'exemple n'est pas completion-$\post$-effective vu le theoreme qui suit}
\begin{proof}
Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
increasing recursive functions from \cite{FMP-ic04} defined as

@@

We start with the case $\Safe=\uparrow \Safe$, hence $\Bad=\downarrow \Bad$.
Since Resilience is undecidable for  $\omega^2$-WSTS with strong compatibility (Theorem \ref{indec WSTS}), we still consider WSTS, and even WBTS, with strong compatibility but by strengthening the assumptions of effectiveness (we now consider the $\post$-ideal-effective hypothesis) and we demonstrate that the three resilience problems are now decidable.

%
\begin{theorem}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a $\post$-ideal-effective WBTS with strong compatibility and a set $\Safe = \uparrow \Safe$.
{\sc Resilience}, {\sc Bounded resilience} 
and {\sc $k$-resilience} are decidable.
\end{theorem}

@@

\begin{proof}
Let us first recall two results in \cite{BFM-ic17} that are stated for WSTS but are also true for WBTS since the proofs rely only on compatibility and not on the property of wqo. Proposition 30 establishes a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ and the runs of its completion $\hat{\mathscr{S}}$. It states that if $x \xrightarrow{k} y$ in $\mathscr{S}$ then for every ideal $I \supseteq \downarrow x$, there exists an ideal $J \supseteq \downarrow y$ such that $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$. Proposition 29 establishes that if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ then for every $y \in J$, there exists $x \in I$ and $y' \geq y$ such that $x \xrightarrow{k'} y'$ in $\mathscr{S}$. Moreover, if $\mathscr{S}$ has transitive compatibility then $k’ \geq k$; if $\mathscr{S}$ has strong compatibility then $k’ = k$.
%

@@

%
\begin{theorem}\label{up-down}
{\sc Resilience}  is decidable for $\pred^*$-basis-effective downward-WSTS with 
$\Safe=\downarrow \Safe$.
\end{theorem}

\begin{proof}
Let $\mathscr{S}$ be a $\pred^*$-basis-effective downward-WSTS with 
$\Safe=\downarrow \Safe$.
By Lemma \ref{downward compatible}, $\pred^*(\Safe)$ is downward-closed; moreover, $\pred^*(\Safe)$ is computable because $\mathscr{S}$ is $\pred^*$-basis-effective.
The resilience problem can be reformulated as $\Bad \cap (S \setminus \pred^*(\Safe)) = \emptyset$.
%\alain{mais comment je calcule $\pred^*(\Safe)$ ?}
Since $\leq$ is ideally-effective, we can compute complementaries and intersections of upward and downward-closed
 subsets.
Hence we can compute $S \setminus \pred^*(\Safe)$ and then the intersection of
$\Bad$
and
$S \setminus \pred^*(\Safe)$,
which are both upward-closed. We may decide whether this intersection is empty, hence
the resilience problem is decidable. \qed
\end{proof}

In the case
of a $\pred^*$-basis-effective WSTS, not necesarly downward compatible,
the above construction
can provide a proof
of non-resilience
i.e. when
$\Bad \cap (S \setminus \downarrow\pred^*(\Safe)) \neq \emptyset$
then
$\Bad \not\subseteq \downarrow\pred^*(\Safe)$
and hence
$\Bad \not\subseteq \pred^*(\Safe)$.
When $\Bad \cap (S \setminus \downarrow\pred^*(\Safe)) = \emptyset$
however
it is not enough to conclude.\\
